---
title: "Cholesteatoma"
date: "06/11/2020"
output: github_document
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(dirname(inputFile),'/README.md')) })

---

<i>Analysis and visualization developed by [Oscar J. Ponce](https://twitter.com/PonceOJ), and project led by [Andrea Paola Solis-Pazmino](https://twitter.com/paosolpaz18)</i>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(dpi=900)
knitr::opts_chunk$set(fig.path = "8 Forestplots/", dev='svg') 


library(webshot)
library(metafor)
library(tidyverse)
library(rmarkdown)
library(magrittr)
library(grid)
library(gt)
library(forestplot)
library(glue)
library(Gmisc)
library(Matrix)


```


## Open Data
>Click any of the following links to view the documents



| Stage  |      File     |  
|:----------:|:-------------:|
| Protocol            | [PROSPERO](https://www.crd.york.ac.uk/prospero/display_record.php?RecordID=184029)|
| Search strategy     | [Ovid](https://github.com/ponceoscarj/Cholesteatoma/blob/main/2%20Search%20Strategy/ovid.csv) and [Scopus](https://github.com/ponceoscarj/Cholesteatoma/blob/main/2%20Search%20Strategy/scopus.csv) | 
| Abstract screening  | [Total Articles = 1024](https://github.com/ponceoscarj/Cholesteatoma/blob/main/3%20Articles%20for%20Abstract%20Screening/AbstractScreening_TotalStudies.txt), [Excluded Articles = 780](https://github.com/ponceoscarj/Cholesteatoma/blob/main/3%20Articles%20for%20Abstract%20Screening/AbstractScreening_ExcludedStudies.txt), **`Included Articles = 247`&#8595;**   |
| Full-text screening |  [Total Articles = 247](https://github.com/ponceoscarj/Cholesteatoma/blob/main/4%20Articles%20for%20Full%20Text%20Screening/FulltextScreening_TotalStudies.txt), [Excluded Articles = 230](https://github.com/ponceoscarj/Cholesteatoma/blob/main/4%20Articles%20for%20Full%20Text%20Screening/FulltextScreening_ExcludedStudies.txt), **`Included Articles = 17`&#8595;**   |
| Data Extraction    | [Total Articles = 17](https://github.com/ponceoscarj/Cholesteatoma/blob/main/5%20Included%20Articles%20for%20Data%20Extraction/IncludedStudies_SRMA.txt) |
| Raw Outcome data    | [Available as `.CSV`](https://github.com/ponceoscarj/Cholesteatoma/blob/main/6%20Extracted%20Data/outcomes.csv)|
| Analysis codes      |    [Script as `R Markdown`](https://github.com/ponceoscarj/Cholesteatoma/blob/main/Cholesteatoma.Rmd)|


```{r cleaning, echo=FALSE, message=FALSE, warning=FALSE}
pao <- read.csv("6 Extracted Data/outcomesV2.csv")

pao$outcome[pao$outcome == 'AC THRESHOLD'] <- 'AC'
pao$outcome[pao$outcome == 'AIR BONE GAP'] <- 'ABG'

pao$outcome[pao$outcome == 'Recurrence'] <- 'recurrence'
pao$outcome[pao$outcome == 'Residual'] <- 'residual'
pao$outcome[pao$outcome == 'RESIDUAL'] <- 'residual'
pao$outcome[pao$outcome == 'Otorrhoea'] <- 'otorrhoea'
pao$outcome[pao$outcome == 'Retraction pocket'] <- 'retraction'

pao$outcome[pao$outcome == 'Recidivism'] <- 'recidivism'

names(pao)[8] <- 'followup'



pao$author <- 
  ifelse(pao$refid==55, paste("Godinho et al., 2005"),
  ifelse(pao$refid==66, paste("Ueda et al., 2001"),
  ifelse(pao$refid==73, paste("Lau et al., 1987"),
  ifelse(pao$refid==84, paste("Gocman et al., 2002"),
  ifelse(pao$refid==100, paste("Dodson et al., 1998"),
  ifelse(pao$refid==205, paste("McGuire et al., 2016"),
  ifelse(pao$refid==238, paste("Sergi et al., 2014"),
  ifelse(pao$refid==270, paste("Drahy et al., 2012"),
  ifelse(pao$refid==340, paste("Chadha et al., 2006"),
  ifelse(pao$refid==343, paste("Shirazi et al., 2006"),
  ifelse(pao$refid==350, paste("Schraff et al., 2005"),
  ifelse(pao$refid==394, paste("Murphy et al., 1998"),
  ifelse(pao$refid==833, paste("Marco-Algarra et al., 1991"),
  ifelse(pao$refid==883, paste("Schmid et al., 1991"),
  ifelse(pao$refid==694, paste("Gamra, 2016"), 
  ifelse(pao$refid==763, paste("Charachon, 1985"), 
  ifelse(pao$refid==839, paste("Moller, 2020"), NA )))))))))))))))))

pao$followup <- ifelse(pao$followup=="", paste("NR"), pao$followup)
pao$followup <- ifelse(pao$followup=="0", paste("No follow-up"), pao$followup)
pao$followup <- ifelse(pao$followup==">12", paste(">1 year"),
                ifelse(pao$followup=="12", paste("1 year"),
                ifelse(pao$followup=="120", paste("10 years"),
                ifelse(pao$followup=="18", paste("1.5 years"),
                ifelse(pao$followup=="22", paste("1.8 years"),
                ifelse(pao$followup=="24", paste("2 years"),         
                ifelse(pao$followup=="3.1 years (mean)", paste("3.1 years"),
                ifelse(pao$followup=="3.2 years", paste("3.2 years"),                       
                ifelse(pao$followup=="3.7 years (mean)", paste("3.7 years"),       
                ifelse(pao$followup=="30", paste("2.5 years"),       
                ifelse(pao$followup=="36", paste("3 years"),
                ifelse(pao$followup=="4", paste("0.3 years"),       
                ifelse(pao$followup=="42", paste("3.5 years"),       
                ifelse(pao$followup=="48", paste("4 years"),
                ifelse(pao$followup=="54", paste("4.5 years"),       
                ifelse(pao$followup=="6", paste("0.5 years"),       
                ifelse(pao$followup=="6.1 year", paste("6.1 years"),       
                ifelse(pao$followup=="60", paste("5 years"),       
                ifelse(pao$followup=="72", paste("6 years"),       
                ifelse(pao$followup=="82", paste("6.8 years"),       
                ifelse(pao$followup=="84", paste("7 years"),
                ifelse(pao$followup=="51 months", paste("4.3 years"),       
                ifelse(pao$followup=="18", paste("18 months"),
                ifelse(pao$followup=="24-56 months", paste("2 to 4.6 years"), pao$followup))))))))))))))))))))))))

```

```{r functions, echo=FALSE, message=FALSE, warning=FALSE}
#functions for relative risk

rr <- function(database){
  db <- database
  db <- escalc(measure="RR", ai=e1, ci=e2, n1i=n1, n2i=n2,data=database)
  db <- db[order(db$yi),]
  db <- summary(db)
  
  db$est <-  exp(db$yi)
  db$ci.lb <- exp(db$ci.lb)
  db$ci.ub <- exp(db$ci.ub)
  
  
  db$rate1 <- paste(db$e1,"/",db$n1)
  db$rate2 <- paste(db$e2,"/",db$n2)
  db$rr <- paste(formatC(db$est, format='f', digits =2)," ",
                 "(",formatC(db$ci.lb, format='f', digits =2),
                 "-",formatC(db$ci.ub, format='f', digits=2),")")  
  
  ma <- rma(db$yi, db$vi, measure='RR', data=db, method='REML', test='knha')
  exp <- predict(ma, transf = transf.exp.int)
  
  db$weights <- weights(ma)
  db$w <- paste(formatC(db$weights, format='f', digits = 0),'%')
  
  list(pre = db, ma = ma, exp = exp)
}



table_rr <- function(analysis, nstudies, follow=FALSE){

  authors <- c("Author", analysis$pre$author, 
      paste("Overall Relative Risk for", analysis$ma$k, "studies"), 
      paste("(Tau^2 = ", (formatC(analysis$ma$tau2, digits=2, format="f")), ", df = ", 
            (analysis$ma$k - analysis$ma$p),
            ", p ", (ifelse(analysis$ma$QEp < 0.001, 
                            paste("< 0.001"),
                            paste("= ", formatC(analysis$ma$QEp, digits=3, format="f")))),
            "; ", "I^2", " = ", (formatC(analysis$ma$I2, digits=1, format="f")), "%)"))
  fu <- c("Follow-up (months)", analysis$pre$followup, NA, NA)
  int <- c("Canal Wall-Up (n/N)",analysis$pre$rate1, paste(sum(analysis$pre$e1), " / ",
                                                      sum(analysis$pre$n1)), NA)
  comp <- c("Canal Wall-Down (n/N)",analysis$pre$rate2, paste(sum(analysis$pre$e2), " / ",
                                                        sum(analysis$pre$n2)), NA)
  rr <- c("Relative Risk (95% CI)", analysis$pre$rr, 
      paste(formatC(analysis$exp$pred, format='f', digits =2), 
            " (",formatC(analysis$exp$ci.lb, format='f', digits=2),
            "-", formatC(analysis$exp$ci.ub, format='f', digits=2), ")"), NA)
  
  weight <- c("Weight(%)", paste(formatC(weights(analysis$ma), format='f', digits = 1),'%'), "100 %", NA)
  
  ifelse(follow==T,
         b <- cbind(authors, int, comp,fu, rr, weight),
         b <- cbind(authors, int, comp, rr, weight))
  
  ifelse(nstudies>1,
         b <- rbind(b[1,], NA, NA, b[2:(nrow(b)-1),], b[nrow(b),]),
         b <- rbind(b[1,], NA, NA, b[2:(nrow(b)-2),], NA))
  
   ifelse(nstudies > 1,
        (c <- structure(list(
          mean = c(rep(NA, 3), analysis$pre$est, exp(analysis$ma$b),NA),
          lower = c(rep(NA, 3), analysis$pre$ci.lb, exp(analysis$ma$ci.lb), NA),
          upper = c(rep(NA, 3), analysis$pre$ci.ub, exp(analysis$ma$ci.ub), NA)),
          .Names = c("mean", "lower", "upper"),
          row.names = c(NA, -1L*nrow(b)),
          class = "data.frame")),
        (c <- structure(list(
          mean = c(rep(NA, 3), analysis$pre$est, NA),
          lower = c(rep(NA, 3), analysis$pre$ci.lb, NA),
          upper = c(rep(NA, 3), analysis$pre$ci.ub, NA)),
          .Names = c("mean", "lower", "upper"),
          row.names = c(NA, -1L*nrow(b)),
          class = "data.frame")))
  c <- as_tibble(c)
  
  list(b = b, c = c)
  
}



plotrr_single_studies <- function(words, numbers,  
                                  xtick, sizes, bolding, aligning, fpPosition,
                                  lines, clip){
  
  forestplot(words,
             graph.pos = fpPosition,
             zero = 1,
             numbers,
             new_page = TRUE,
             colgap = unit(5, "mm"),
             hrzl_lines = lines,
             lineheight=unit(0.7,'cm'),
             boxsize = sizes,
             line.margin = 2,
             is.summary = bolding,
             align = aligning,
             ci.vertices = TRUE,
             txt_gp = fpTxtGp(label =gpar (cex=0.8), 
                              ticks = gpar(cex = 0.8, fontface="bold"),
                              summary = gpar(cex = 0.8),
                              xlab = gpar(cex=0.8)),
             xticks = xtick,
             xlog=TRUE,
             clip = clip,
             grid = xtick,
             lwd.xaxis = 1,
             lwd.ci = 2.2,
             lwd.zero = 1.5,
             graphwidth = unit(10,"cm"),
             col=fpColors(box="black",line="grey", axes="grey20", summary="black", zero = 'dodgerblue4'))
}

#functions for md


md <- function(database){
  db <- database
  db <- escalc(measure="MD", 
               m1i=mean_post1, m2i = mean_post2, 
               sd1i  = sd_post1, sd2i = sd_post2, 
               n1i = n1, n2i= n2, data=database)
  db$vi <- ifelse(is.na(db$vi), 
               ((db$mdul_post-db$mdll_post)/((2*abs(qt(0.05/2, db$total-1)))^2)), db$vi)
  db <- db[order(db$yi),]
  db <- summary(db)
  
  db$md <- paste(formatC(db$yi, format='f', digits =2)," ",
                 "(",formatC(db$ci.lb, format='f', digits =2),
                 ",",formatC(db$ci.ub, format='f', digits=2),")") 
  db$postmean1 <- paste((ifelse(is.na(db$mean_post1), paste("NR"), formatC(db$mean_post1, format='f', digits=1))), '(', (ifelse(is.na(db$sd_post1), paste("NR"),formatC(db$sd_post1, format='f', digits=1))),')')
  db$postmean2 <- paste((ifelse(is.na(db$mean_post2), paste("NR"), formatC(db$mean_post2, format='f', digits=1))),'(', (ifelse(is.na(db$sd_post2), paste("NR"),formatC(db$sd_post2, format='f', digits=1))),')')
  
  ma <- rma(yi, vi, measure='MD', data=db, method='REML', test='knha')
  
  db$weights <- weights(ma)
  db$w <- paste(formatC(db$weights, format='f', digits = 0),'%')
  
  list(pre = db, ma = ma)
}



table_md <- function(analysis, nstudies, int, comp, outcome, col,
                     follow=FALSE){

      authors <- c("Author", analysis$pre$author, 
                    paste("Overall for ", analysis$ma$k, "studies"), 
                    paste("(Tau^2 = ", (formatC(analysis$ma$tau2, digits=2, format="f")), ", df = ", 
                    (analysis$ma$k - analysis$ma$p),
                    ", p ", (ifelse(analysis$ma$QEp < 0.001, 
                                    paste("< 0.001"),
                                    paste("= ", formatC(analysis$ma$QEp, digits=3, format="f")))),
                    "; ", "I^2", " = ", (formatC(analysis$ma$I2, digits=1, format="f")), "%)"))
      comparison <- c("Comparison", analysis$pre$comparison, NA, NA)
      fu         <- c("Follow-up (months)", analysis$pre$followup, NA, NA)
      int_pop    <- c(paste(int), analysis$pre$n1, sum(analysis$pre$n1), NA)
      int_cont   <- c(paste(col), analysis$pre$postmean1, NA, NA)
      comp_pop   <- c(paste(comp), analysis$pre$n2,sum(analysis$pre$n2), NA)
      comp_cont  <- c(paste(col), analysis$pre$postmean2, NA, NA)
      smd        <- c(paste0("Mean Difference (95% CI)"),
                      analysis$pre$md, 
                      paste(formatC(analysis$ma$b, format='f', digits =2), 
                            " (",formatC(analysis$ma$ci.lb, format='f', digits=2),
                            ",", formatC(analysis$ma$ci.ub, format='f', digits=2), ")"), NA)
      weight     <- c("Weight (%)", analysis$pre$w, "100 %", NA)


  ifelse(follow==T,
         b <- cbind(authors, int_pop, int_cont, comp_pop, comp_cont, fu,  smd, weight),
         b <- cbind(authors, int_pop, int_cont, comp_pop, comp_cont, smd, weight))
  
  ifelse(nstudies>1,
         b <- rbind(b[1,], NA, NA, b[2:(nrow(b)-1),], b[nrow(b),]),
         b <- rbind(b[1,], NA, NA, b[2:(nrow(b)-2),], NA))
  
  ifelse(nstudies > 1,
        (c <- structure(list(
          mean = c(rep(NA, 3), analysis$pre$yi, analysis$ma$b,NA),
          lower = c(rep(NA, 3), analysis$pre$ci.lb, analysis$ma$ci.lb, NA),
          upper = c(rep(NA, 3), analysis$pre$ci.ub, analysis$ma$ci.ub, NA)),
          .Names = c("mean", "lower", "upper"),
          row.names = c(NA, -1L*nrow(b)),
          class = "data.frame")),
        (c <- structure(list(
          mean = c(rep(NA, 3), analysis$pre$yi, NA),
          lower = c(rep(NA, 3), analysis$pre$ci.lb, NA),
          upper = c(rep(NA, 3), analysis$pre$ci.ub, NA)),
          .Names = c("mean", "lower", "upper"),
          row.names = c(NA, -1L*nrow(b)),
          class = "data.frame")))

  c <- as_tibble(c)
  
  list(b = b, c = c)
} 





plotmd_single_studies <- function(words, numbers,  
                                  xtick, sizes, bolding, aligning, fpPosition,
                                  lines) {
  (forestplot(words,
             graph.pos = fpPosition,
             zero = 0,
             numbers,
             new_page = TRUE,
             colgap = unit(5, "mm"),
             hrzl_lines = lines,
             lineheight=unit(0.7,'cm'),
             boxsize = sizes,
             line.margin = 2,
             is.summary = bolding,
             align = aligning,
             ci.vertices = TRUE,
             txt_gp = fpTxtGp(label =gpar (cex=0.9), 
                              ticks = gpar(cex = 0.9, fontface="bold"),
                              summary = gpar(cex = 0.9, fontsize=13),
                              xlab = gpar(cex=0.9)),
             xticks = xtick,
             xlog=FALSE,
             clip = c(0,1),
             grid = xtick,
             lwd.xaxis = 1,
             lwd.ci = 2.2,
             lwd.zero = 1.5,
             graphwidth = unit(10,"cm"),
             col=fpColors(box=c("black"),line="grey", zero = 'dodgerblue4', axes="grey20", summary='black')))
}


```


## Forest plots

### Figure 2 

<details><summary><b>Show figure -</b> Differences in <b>Air Bone Gap (db)</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r abg, fig.height=3, fig.width=19, echo=FALSE, message=FALSE, warning=FALSE}
abg <- subset(pao, outcome=='ABG')
abg <- abg[!(abg$refid==55),]


md1 <- md(abg)
tmd1 <- table_md(analysis = md1, nstudies = 4, int = "Canal Wall-Up (n)",
                    comp = "Canal Wall-Down (n)", outcome = "Air Bone Gap", col = "Mean (SD)",
                 follow = TRUE)

plotmd_single_studies(words = tmd1$b,
                      numbers = tmd1$c, 
                      fpPosition = ncol(tmd1$b)-1,
                      xtick = c(-15,-5,5,15),
                      sizes = c(rep(NA,3), 0.006*(md1$pre$weights+50),1,NA),
                      bolding = c(T, rep(F,6), T, T),
                      aligning = c("l", rep("c",4), "l", "l"),
                      lines = list('3' = gpar (lwd=1, columns=c(1:(ncol(tmd1$b)+1)), col="black")))
                    
```


</details>


### Figure 3 
<details><summary><b>Show figure -</b> Difference in <b>Air Conduction (Hz)</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r ac, fig.height=3, fig.width=16, echo=FALSE, message=FALSE, warning=FALSE}
ac <- subset(pao, outcome=='AC')

md2 <- md(ac)
tmd2 <- table_md(analysis = md2, nstudies = 1, int = "Canal Wall-Up (n)",
                    comp = "Canal Wall-Down (n)", outcome = "Air Bone Gap",col = "Mean (SD)",
                 follow = TRUE)

plotmd_single_studies(words = tmd2$b,
                      numbers = tmd2$c, 
                      fpPosition = ncol(tmd2$b)-1,
                      xtick = c(-25,-15,-5,5,15, 25),
                      sizes = c(rep(NA,3),1,NA),
                      bolding = c(T, rep(F,2), F, F),
                      aligning = c("l", rep("c",4), "l", "l"),
                      lines = list('3' = gpar (lwd=1, columns=c(1:(ncol(tmd2$b)+1)), col="black")))


```

</details>



### Figure 4
<details><summary><b>Show figure -</b> Risk of <b>Recurrence</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r recurrence, fig.height=4, fig.width=15, echo=FALSE, message=FALSE, warning=FALSE}
#recurrence longest followup
rcr <- subset(pao, outcome=='recurrence')
rownames(rcr) <- NULL

#keeping longest follow-up for 763 and only "stage-one" group which is = to recurrence
rcr <- rcr[-c(7,8,10),]

recu <- rr(rcr)

trecu <- table_rr(analysis = recu,nstudies = 12,follow = TRUE)

plotrr_single_studies(words = trecu$b,
                      numbers = trecu$c, 
                      fpPosition =  ncol(trecu$b)-1,
                      xtick = c(0.1, 0.2, 0.5, 2, 4, 8, 16,32),
                      clip = c(0.25,8),
                      sizes = c(rep(NA,3),0.006*(recu$pre$weights+50),1),
                      bolding = c(T,rep(F,11), T,T),
                      aligning = c("l", rep("c",2), "l", "l"),
                      lines = list('3' = gpar (lwd=1, columns=c(1:(ncol(trecu$b)+1)), col="black")))

                       

```

</details>




### Figure 5
<details><summary><b>Show figure -</b> Risk of <b>Residual Tissue</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r residual, fig.height=4, fig.width=15, echo=FALSE, message=FALSE, warning=FALSE}
#residual 
res <- subset(pao, outcome=='residual')

residual <- rr(res)

tresidual <- table_rr(analysis = residual ,nstudies = 10,follow = TRUE)

plotrr_single_studies(words = tresidual$b,
                      numbers = tresidual$c, 
                      fpPosition =  ncol(tresidual$b)-1,
                      xtick = c(0.1, 0.2, 0.5, 2, 4, 8, 16),
                      clip = c(0.25,8),
                      sizes = c(rep(NA,3),0.006*(residual$pre$weights+50),1),
                      bolding = c(T,rep(F,11), T,T),
                      aligning = c("l", rep("c",2), "l", "l"),
                      lines = list('3' = gpar (lwd=1, columns=c(1:(ncol(tresidual$b)+1)), col="black")))

```

</details>



### Figure 6 
<details><summary><b>Show figure -</b> Risk of <b>Recidivism</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r recidivism, fig.height=3.5, fig.width=15, echo=FALSE, message=FALSE, warning=FALSE}
#meta analysis recidivism longest followup
prerec <- subset(pao, outcome=='recidivism' | outcome=='recurrence' | outcome=='residual')

rownames(prerec) <- NULL

prerec <- prerec[c(9,11,13:15,17,19:21,22,24,27,31,34,39,41),]
prerec <- prerec[,c(1,46,2:45)]

prerec <- subset(prerec, outcome=='recidivism')

rec <- prerec

reci <- rr(rec)


treci <- table_rr(analysis = reci ,nstudies = 10,follow = TRUE)

plotrr_single_studies(words = treci$b,
                      numbers = treci$c, 
                      fpPosition =  ncol(treci$b)-1,
                      xtick = c(0.25, 0.5, 2, 4, 8),
                      clip = c(0.25,8),
                      sizes = c(rep(NA,3),0.006*(reci$pre$weights+50),1),
                      bolding = c(T,rep(F,9), T,T),
                      aligning = c("l", rep("c",2), "l", "l"),
                      lines = list('3' = gpar (lwd=1, columns=c(1:(ncol(treci$b)+1)), col="black")))

```


</details>
