---
title: "Cholesteatoma"
author: "Oscar J. Ponce & Paola Solis-Pazmino"
date: "06/11/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(dpi=900)
knitr::opts_chunk$set(dev='svg')

library(webshot)
library(metafor)
library(tidyverse)
library(rmarkdown)
library(magrittr)
library(grid)
library(gt)
library(forestplot)
library(glue)
library(Gmisc)
library(Matrix)


```


## Open Data
>Click any of the following links to view the documents



| Stage  |      File     |  
|:----------:|:-------------:|
| Protocol            | [PROSPERO](https://www.crd.york.ac.uk/prospero/display_record.php?RecordID=184029)|
| Search strategy     | [Ovid](https://github.com/ponceoscarj/Cholesteatoma/blob/main/2%20Search%20Strategy/ovid.csv) and [Scopus](https://github.com/ponceoscarj/Cholesteatoma/blob/main/2%20Search%20Strategy/scopus.csv) | 
| Abstract screening  | [Total Articles = 1024](https://github.com/ponceoscarj/Cholesteatoma/blob/main/3%20Articles%20for%20Abstract%20Screening/AbstractScreening_TotalStudies.txt), [Excluded Articles = 780](https://github.com/ponceoscarj/Cholesteatoma/blob/main/3%20Articles%20for%20Abstract%20Screening/AbstractScreening_ExcludedStudies.txt), **`Included Articles = 247`&#8595;**   |
| Full-text screening |  [Total Articles = 247](https://github.com/ponceoscarj/Cholesteatoma/blob/main/4%20Articles%20for%20Full%20Text%20Screening/FulltextScreening_TotalStudies.txt), [Excluded Articles = 230](https://github.com/ponceoscarj/Cholesteatoma/blob/main/4%20Articles%20for%20Full%20Text%20Screening/FulltextScreening_ExcludedStudies.txt), **`Included Articles = 17`&#8595;**   |
| Data Extraction    | [Total Articles = 17](https://github.com/ponceoscarj/Cholesteatoma/blob/main/5%20Included%20Articles%20for%20Data%20Extraction/IncludedStudies_SRMA.txt) |
| Raw Outcome data    | [Available as `.CSV`](https://github.com/ponceoscarj/Cholesteatoma/blob/main/6%20Extracted%20Data/outcomes.csv)|
| Analysis codes      |    [Script as `R Markdown`](https://github.com/ponceoscarj/Cholesteatoma/blob/main/Cholesteatoma.Rmd)|


```{r cleaning, echo=FALSE, message=FALSE, warning=FALSE}
pao <- read.csv("6 Extracted Data/outcomes.csv")

pao$outcome[pao$outcome == 'AC THRESHOLD'] <- 'AC'
pao$outcome[pao$outcome == 'AIR BONE GAP'] <- 'ABG'

pao$outcome[pao$outcome == 'Recurrence'] <- 'recurrence'
pao$outcome[pao$outcome == 'Residual'] <- 'residual'
pao$outcome[pao$outcome == 'Otorrhoea'] <- 'otorrhoea'
pao$outcome[pao$outcome == 'Retraction pocket'] <- 'retraction'

pao$variable <- ifelse(is.na(pao[18] | pao[20] | pao[22] | pao[28] |
                               pao[34] | pao[40]), paste('dic'), paste('cont'))



pao$author <- 
  ifelse(pao$refid==55, paste("Godinho et al., 2005"),
  ifelse(pao$refid==66, paste("Ueda et al., 2001"),
  ifelse(pao$refid==73, paste("Lau et al., 1987"),
  ifelse(pao$refid==84, paste("Gocman et al., 2002"),
  ifelse(pao$refid==100, paste("Dodson et al., 1998"),
  ifelse(pao$refid==205, paste("McGuire et al., 2016"),
  ifelse(pao$refid==238, paste("Sergi et al., 2014"),
  ifelse(pao$refid==270, paste("Drahy et al., 2012"),
  ifelse(pao$refid==340, paste("Chadha et al., 2006"),
  ifelse(pao$refid==343, paste("Shirazi et al., 2006"),
  ifelse(pao$refid==350, paste("Scott et al., 2005"),
  ifelse(pao$refid==394, paste("Murphy et al., 1998"),
  ifelse(pao$refid==833, paste("Marco-Algarra et al., 1991"),
  ifelse(pao$refid==883, paste("Schmid et al., 1991"),
  ifelse(pao$refid==694, paste("Gamra, 2016"), 
  ifelse(pao$refid==763, paste("Charachon, 1985"), 
  ifelse(pao$refid==839, paste("Ronde, 2020"), NA )))))))))))))))))
```

```{r functions, echo=FALSE, message=FALSE, warning=FALSE}
#functions for relative risk

rr <- function(database){
  db <- database
  db <- escalc(measure="RR", ai=e1, ci=e2, n1i=n1, n2i=n2,data=database)
  db <- db[order(db$yi),]
  db <- summary(db)
  
  db$est <-  exp(db$yi)
  db$ci.lb <- exp(db$ci.lb)
  db$ci.ub <- exp(db$ci.ub)
  
  
  db$rate1 <- paste(db$e1,"/",db$n1)
  db$rate2 <- paste(db$e2,"/",db$n2)
  db$rr <- paste(formatC(db$est, format='f', digits =2)," ",
                 "(",formatC(db$ci.lb, format='f', digits =2),
                 "-",formatC(db$ci.ub, format='f', digits=2),")")  
  
  ma <- rma(db$yi, db$vi, measure='RR', data=db, method='REML')
  exp <- predict(ma, transf = transf.exp.int)
  
  list(pre = db, ma = ma, exp = exp)
}



table_rr <- function(e1, n1, e2, n2, analysis){
  b <- cbind( 
    c("Author", analysis$pre$author, 
      paste("Overall Relative Risk for", analysis$ma$k, "studies","\n", 
            "(Tau^2 = ", (formatC(analysis$ma$tau2, digits=2, format="f")), ", df = ", 
            (analysis$ma$k - analysis$ma$p),
            ", p ", (ifelse(analysis$ma$QEp < 0.001, 
                            paste("< 0.001"),
                            paste("= ", formatC(analysis$ma$QEp, digits=3, format="f")))),
            "; ", "I^2", " = ", (formatC(analysis$ma$I2, digits=1, format="f")), "%)")),
    c("Canal wall up (n/N)",analysis$pre$rate1, paste(sum(e1), " / ",
                                                      sum(n1))),
    c("Canal wall down (n/N)",analysis$pre$rate2, paste(sum(e2), " / ",
                                                        sum(n2))),
    c("Relative Risk (95% CI)", analysis$pre$rr, 
      paste(formatC(analysis$exp$pred, format='f', digits =2), 
            " (",formatC(analysis$exp$ci.lb, format='f', digits=2),
            "-", formatC(analysis$exp$ci.ub, format='f', digits=2), ")")),
    c("Weight(%)", paste(formatC(weights(analysis$ma), format='f', digits = 1),'%'), NA))
  
  b <- rbind(b, NA)
  b <- as_tibble(b, .name_repair = "unique")
  b <- b %>% add_row(.before = (nrow(b)-1))
  
  c <- structure(list(
    mean = c(NA,  analysis$pre$est, NA, analysis$exp$pred, NA),
    lower = c(NA,   analysis$pre$ci.lb, NA, analysis$exp$ci.lb, NA),
    upper = c(NA,   analysis$pre$ci.ub, NA, analysis$exp$ci.ub, NA)),
    .Names = c("mean", "lower", "upper"),
    row.names = c(NA, (-1L*nrow(b))),
    class = "data.frame")
  
  
  list(b = b, c = c)
  
}



plotrr_single_studies <- function(words, numbers, sizebox, box_ma_results, overall_box, xcoord){
  
  box_size <- (sizebox*(weights(box_ma_results)+50))
  
  forestplot(words,
             graph.pos = ncol(words)-1,
             zero = 1,
             numbers,
             new_page = TRUE,
             colgap = unit(5, "mm"),
             hrzl_lines = list("2" = gpar (lwd=1, columns=c(1:6), col="black")),
             lineheight=unit(0.7,'cm'),
             boxsize = c(NA, box_size, NA, overall_box, NA),
             line.margin = 2,
             is.summary = c(T, rep(F, nrow(words)-3), T, F),
             align = c("l","c", "c", "l", "l"),
             ci.vertices = TRUE,
             txt_gp = fpTxtGp(label =gpar (cex=0.8), 
                              ticks = gpar(cex = 0.8, fontface="bold"),
                              summary = gpar(cex = 0.8),
                              xlab = gpar(cex=0.8)),
             xticks = xcoord,
             xlog=TRUE,
             clip = c(0.2, 32),
             grid = xcoord,
             lwd.xaxis = 1,
             lwd.ci = 2.2,
             graphwidth = unit(10,"cm"),
             col=fpColors(box="black",line="grey", axes="grey20", summary="black"))
}

#functions for md


md <- function(database){
  db <- database
  db <- escalc(measure="MD", 
               m1i=mean_post1, m2i = mean_post2, 
               sd1i  = sd_post1, sd2i = sd_post2, 
               n1i = n1, n2i= n2, data=database)
  db$vi <- ifelse(is.na(db$vi), 
               ((db$mdul_post-db$mdll_post)/((2*abs(qt(0.05/2, db$total-1)))^2)), db$vi)
  db <- db[order(db$yi),]
  db <- summary(db)
  
  db$md <- paste(formatC(db$yi, format='f', digits =2)," ",
                 "(",formatC(db$ci.lb, format='f', digits =2),
                 ",",formatC(db$ci.ub, format='f', digits=2),")")  
  
  ma <- rma(yi, vi, measure='MD', data=db, method='REML')

  list(pre = db, ma = ma)
}


table_md <- function(n1, n2, analysis){
  b <- cbind( 
    c("Author", analysis$pre$author, 
      paste("Overall Mean Difference for", analysis$ma$k, "studies","\n", 
            "(Tau^2 = ", (formatC(analysis$ma$tau2, digits=2, format="f")), ", df = ", 
            (analysis$ma$k - analysis$ma$p),
            ", p ", (ifelse(analysis$ma$QEp < 0.001, 
                            paste("< 0.001"),
                            paste("= ", formatC(analysis$ma$QEp, digits=3, format="f")))),
            "; ", "I^2", " = ", (formatC(analysis$ma$I2, digits=1, format="f")), "%)")),
    c("Canal wall up (N)",analysis$pre$n1, sum(n1)),
    c("Canal wall down (N)",analysis$pre$n2,sum(n2)),
    c("Mean Difference (95% CI)", analysis$pre$md, 
      paste(formatC(analysis$ma$b, format='f', digits =2), 
            " (",formatC(analysis$ma$ci.lb, format='f', digits=2),
            ",", formatC(analysis$ma$ci.ub, format='f', digits=2), ")")),
    c("Weight(%)", paste(formatC(weights(analysis$ma), format='f', digits = 1),'%'), NA))
  
  b <- rbind(b, NA)
  b <- as_tibble(b, .name_repair = "unique")
  b <- b %>% add_row(.before = (nrow(b)-1))
  
  c <- structure(list(
    mean = c(NA,  analysis$pre$yi, NA, analysis$ma$b, NA),
    lower = c(NA,   analysis$pre$ci.lb, NA, analysis$ma$ci.lb, NA),
    upper = c(NA,   analysis$pre$ci.ub, NA, analysis$ma$ci.ub, NA)),
    .Names = c("mean", "lower", "upper"),
    row.names = c(NA, (-1L*nrow(b))),
    class = "data.frame")
  
  
  list(b = b, c = c)
  
}



plotmd_single_studies <- function(words, numbers, sizebox, box_ma_results, overall_box, xcoord){
  
  box_size <- (sizebox*(weights(box_ma_results)+50))
  
  forestplot(words,
             graph.pos = ncol(words)-1,
             zero = 0,
             numbers,
             new_page = TRUE,
             colgap = unit(5, "mm"),
             hrzl_lines = list("2" = gpar (lwd=1, columns=c(1:6), col="black")),
             lineheight=unit(0.7,'cm'),
             boxsize = c(NA, box_size, NA, overall_box, NA),
             line.margin = 2,
             is.summary = c(T, rep(F, nrow(words)-3), T, F),
             align = c("l","c", "c", "l", "l"),
             ci.vertices = TRUE,
             txt_gp = fpTxtGp(label =gpar (cex=0.8), 
                              ticks = gpar(cex = 0.8, fontface="bold"),
                              summary = gpar(cex = 0.8),
                              xlab = gpar(cex=0.8)),
             xticks = xcoord,
             xlog=FALSE,
             clip = c(0.2, 32),
             grid = xcoord,
             lwd.xaxis = 1,
             lwd.ci = 2.2,
             graphwidth = unit(10,"cm"),
             col=fpColors(box="black",line="grey", axes="grey20", summary="black"))
}


```


## Forest plots

### Figure 2 
<details><summary><b>Show figure -</b> Differences in <b>Air Bone Gap (db)</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r abg, fig.height=3, fig.width=13, echo=FALSE, message=FALSE, warning=FALSE}
abg <- subset(pao, variable=='cont' & outcome=='ABG')
abg <- abg[!(abg$refid==55),]

###evaluation of correlation coefficients
abg$corre <- (abg$sd_pre1^2 + abg$sd_post1^2 - abg$sd_ch1^2)/(2*abg$sd_pre1*abg$sd_post1)
abg$corrc <- (abg$sd_pre2^2 + abg$sd_post2^2 - abg$sd_ch2^2)/(2*abg$sd_pre2*abg$sd_post2)


md1 <- md(abg)
tmd1 <- table_md(n1 = abg$n1, n2 = abg$n2,
                 analysis = md1)

plotmd_single_studies(words = tmd1$b,
                      numbers = tmd1$c, 
                      sizebox = 0.008,
                      box_ma_results = md1$ma,
                      overall_box = 1, xcoord = c(-15, -10, -5, 0, 5, 10, 15) )

```


</details>


### Figure 3 
<details><summary><b>Show figure -</b> Difference in <b>Air Conduction (Hz)</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r ac, fig.height=3, fig.width=11, echo=FALSE, message=FALSE, warning=FALSE}
ac <- subset(pao, variable=='cont' & outcome=='AC')

md2 <- md(ac)
tmd2 <- table_md(n1 = ac$n1, n2 = ac$n2,
                 analysis = md2)

tmd2$b <- tmd2$b[c(1:3),]
tmd2$c <- tmd2$c[c(1:3),]

for (i in 1:length(tmd2)){
  tmd2[[i]] <- tmd2[[i]] %>% add_row(.before = 2)
}



forestplot(tmd2$b,
             graph.pos = ncol(tmd2$b)-1,
             zero = 0,
             tmd2$c,
             new_page = TRUE,
             colgap = unit(5, "mm"),
             hrzl_lines = list("2" = gpar (lwd=1, columns=c(1:6), col="black")),
             lineheight=unit(0.7,'cm'),
             boxsize = c(NA, NA, 1, NA),
             line.margin = 2,
             is.summary = c(T, F, T, F),
             align = c("l","c", "c", "l", "l"),
             ci.vertices = TRUE,
             txt_gp = fpTxtGp(label =gpar (cex=0.8), 
                              ticks = gpar(cex = 0.8, fontface="bold"),
                              summary = gpar(cex = 0.8),
                              xlab = gpar(cex=0.8)),
             xticks = c(-20,-10,0, 10),
             xlog=FALSE,
             clip = c(0.2, 32),
             grid = c(-20,-10,0, 10),
             lwd.xaxis = 1,
             lwd.ci = 2.2,
             graphwidth = unit(10,"cm"),
             col=fpColors(box="black",line="grey", axes="grey20", summary="black"))
  


```

</details>



### Figure 4
<details><summary><b>Show figure -</b> Risk of <b>Recurrence</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r recurrence, fig.height=4, fig.width=13, echo=FALSE, message=FALSE, warning=FALSE}
#recurrence longest followup
rcr <- subset(pao, variable=='dic' & outcome=='recurrence')

rcr<-rcr[!(rcr$refid==55 & rcr$fu!='54'),]
rcr<-rcr[!(rcr$refid==839 & rcr$fu!='60'),]


recu <- rr(rcr)

trecu <- table_rr(e1 = rcr$e1, n1 = rcr$n1, 
                  e2 = rcr$e2, n2 = rcr$n2, 
                  analysis = recu)

plotrr_single_studies(words = trecu$b,
                      numbers = trecu$c, 
                      sizebox = 0.008,
                      box_ma_results = recu$ma,
                      overall_box = 1, xcoord = c(0.06, 0.25,  1, 4, 16))

```

</details>




### Figure 5
<details><summary><b>Show figure -</b> Risk of <b>Residual Tissue</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r residual, fig.height=4, fig.width=13, echo=FALSE, message=FALSE, warning=FALSE}
#residual 
res <- subset(pao, variable=='dic' & outcome=='residual')

residual <- rr(res)

tresidual <- table_rr(e1 = res$e1, n1 = res$n1, 
                  e2 = res$e2, n2 = res$n2, 
                  analysis = residual)

plotrr_single_studies(words = tresidual$b,
                      numbers = tresidual$c, 
                      sizebox = 0.006,
                      box_ma_results = residual$ma,
                      overall_box = 1, xcoord = c(0.06, 0.25 , 1, 4,  16))
```

</details>



### Figure 6 
<details><summary><b>Show figure -</b> Risk of <b>Recidivism</b> after <b>Canal Wall Up</b> versus <b>Canal Wall Down</b></summary>

```{r recidivism, fig.height=6, fig.width=13, echo=FALSE, message=FALSE, warning=FALSE}
#meta analysis recidivism longest followup
prerec <- subset(pao, outcome=='recidivism' | outcome=='recurrence' | outcome=='residual')

prerec<-prerec[!(prerec$refid==55 & prerec$fu!='54'),]
prerec<-prerec[!(prerec$refid==66 & prerec$outcom=='residual'),]
prerec<-prerec[!(prerec$refid==73 & prerec$outcom=='residual'),]
prerec<-prerec[!(prerec$refid==100 & prerec$outcom=='residual'),]
prerec<-prerec[!(prerec$refid==238 & prerec$outcom=='residual'),]
prerec<-prerec[!(prerec$refid==343 & prerec$outcom=='residual'),]
prerec<-prerec[!(prerec$refid==694 & prerec$outcom=='residual'),]
prerec<-prerec[!(prerec$refid==833 & prerec$fu!='60'),]
prerec<-prerec[!(prerec$refid==839 & prerec$fu!='60'),]
prerec<-prerec[!(prerec$refid==883 & prerec$fu!='residual'),]

rec <- prerec

reci <- rr(rec)

treci <- table_rr(e1 = rec$e1, n1 = rec$n1, 
                      e2 = rec$e2, n2 = rec$n2, 
                      analysis = reci)

plotrr_single_studies(words = treci$b,
                      numbers = treci$c, 
                      sizebox = 0.008,
                      box_ma_results = reci$ma,
                      overall_box = 1, xcoord = c(0.06, 0.25, 1, 4, 16))

```


</details>
